Combining information from photometric and spectroscopic data:\\ Probabilistic spectrophotometric inference

### 2015-07-08 tex/figures.tex
more figure code
tweak text in S2, and elsewhere, switch from A_v to tau_V

\begin{figure*}[h!]
%\includegraphics[width=\textwidth]{figures/calibration_posterior.pdf}
\caption{Joint Posterior PDFs for the calibration parameters inferred
from the combination of \emph{uncalibrated} spectroscopy and all the
photometric data points.
\label{fig:cal_posterior}}
\end{figure*}

@@

\begin{figure*}[h!]
\includegraphics[width=\textwidth]{figures/vary_nphot.pdf}
\caption{The value of more photometry. Marginalized posterior PDFs
obtained from mock spectra and photometry, for different numbers of
photometric bands.  The pdfs $g$-band only ({\it red}), $g$- and $i-$
band ({\it blue}), and the combined SDSS $griz$ and 2MASS $JHK_s$ set of
bands ({\it orange}).
\label{fig:vary_phot}}
\end{figure*}

@@

\begin{figure*}[h!]
%\includegraphics[width=\textwidth]{figures/real_phys_posterior.png}
\caption{Inferred parameters for \excluster.  We show the inference from the
  calibrated spectrum ({\it red}), from the uncalibrated spectrum
  ({\it blue}), and from the CMD ({\it black}).
\label{fig:real_parameters}}
\end{figure*}

>>>>

\begin{figure*}[h!]
\includegraphics[width=\textwidth]{figures/calibration_post.pdf}
\caption{Joint Posterior PDFs for the calibration parameters inferred
from the combination of \emph{uncalibrated} spectroscopy and all the
photometric data points.
\label{fig:cal_posterior}}
\end{figure*}

@@

\begin{figure*}[h!]
\includegraphics[width=\textwidth]{figures/vary_nphot.pdf}
\caption{The value of more photometry in combination with optical
spectroscopy. Marginalized posterior PDFs obtained from mock spectra
and photometry, for different numbers of photometric bands. The
medians of the posterior PDFs are shown as connected black circles,
the 16th and 84th percentile of the posteriors are indicated by the
gray shaded region.
\label{fig:vary_phot}}
\end{figure*}

@@

\begin{figure*}[h!]
\includegraphics[width=\textwidth]{figures/real_post.pdf}
\caption{Inferred parameters for \excluster.  We show the inference from the
  calibrated spectrum ({\it green}), from the uncalibrated spectrum
  ({\it orange}), and from the CMD ({\it black}).
\label{fig:real_parameters}}
\end{figure*}

### 2015-07-08 tex/ms.tex
tweak text in S2, and elsewhere, switch from A_v to tau_V

A model is a parameterized likelihood function, with prior PDFs
on the nuisance parameters (at least).  There is a likelihood for
the photometry and for the spectroscopy...  Each of these penalizes
the difference between a column vector of data and a column vector of
model expectation value.

@@

If you don't believe the reported data noise variances $\sigma_n^2$,
it often makes sense to add in quadrature a ``jitter'' variance $s^2$.
This is a simple modification of the chi-squared log-likelihood:
\begin{eqnarray}\label{eq:photometricLF}
\ln p(d\given\theta,s^2) &=& -\frac{1}{2}\,\sum_{n=1}^N
                             \left[\frac{[d_n - \mu_n(\theta)]^2}
                             {[\sigma_n^2 + s^2]} + 
                             \ln(2\pi\,[\sigma_n^2 + s^2]) \right]
\quad .
\end{eqnarray}
This modification shows the necessity of including the log-variance
term.
This---equation~(\ref{eq:photometricLF})---is in fact the
log-likelihood function we are going to use for the photometric
measurements in what follows.  For photometric data this ``jitter'' is
often best described as a fractional error, $s^2 \equiv
\hat{s}^2\mu(\theta)^2$, corresponding for example to an additional
magnitude zero-point uncertainty.  In this formulation, if $\theta$
are the parameters of interest, the jitter variance $\hat{s} =
s/\mu(\theta)$ is a nuisance noise-model parameter representing
additonal fractional uncertainty in the flux.  The Gaussian Process
will be a generalization of this modification.

{\color{blue} BDJ This paragraph should explain why we are using the
GP, not just how it works.}
For the spectroscopic measurements we are going to account for the
spectrophotometric calibration in our likelihood function.  We take
this calibration-vector that modifies the mean spectroscopic model to
produce the observed spectrum to be multiplicative with respect to the
mean model.  The multiplicative calibration vector can be explicitly
modeled, for example using polynomials.  However, for complex
calibration vectors over large wavelength ranges, the number of
parameters required to model the calibration can become quite large
(e.g. very high order polynomials) and the correct parameterization
may be unknown.  It is therefore desirable to introduce a
non-parameteric model of the calibration.  We do this using a Gaussian
Process: 
BDJ: prior over functions, analytically marginalizable, covariant
fractional uncertainties the same as multiplying by a function
(covariant absolute uncertainties the same as adding a function.)
Using a polynomial can result in degeneracies between model and
polynomial parameters, whereas GP only reduces the penalty for
residuals that {\it can't} be fit by the mean model.

The calibration-vector and Gaussian-Process modified log-likelihood is
\begin{eqnarray}\label{eq:spectroscopicLF_v2}
\ln p(d\given\theta,\alpha) &=&
                                -\frac{1}{2}\,\left[\transpose{\Delta}\,C^{-1}\,\Delta
                                + \ln([2\pi]^N\,\det{C}) \right]
\\
\Delta &\equiv& d - \mu(\theta) \ast f(\alpha) 
\\
C_{nn'} &\equiv& [\sigma_n^2 + (\hat{s} \, \mu_n)^2]\,\delta_{nn'} +
 K_\alpha(\lambda_n - \lambda_{n'}) 
\quad ,
\end{eqnarray}
where $\ast$ denotes element-wise multiplication,
$\theta$ are the physical parameters,
$\alpha$ is a blob of nuisance parameters
(which determine the jitter, the calibration vector and the kernel function),
$\Delta$ is a residual column vector,
$f(\alpha)$ is a calibration column vector,
$C$ is the Gaussian Process covariance matrix,
made up of $N^2$ values $C_{nn'}$,
$K_\alpha$ is a kernel function that sets the wavelength scale and
amplitude of the residual calibration issues,
and each $\lambda_n$ is the wavelength of spectral pixel $n$.
The calibration vector $f(\alpha)$ is going to take care of the
large-scale calibration residual; the Gaussian Process is going to
pick up smaller-scale residuals away from \emph{that}.
The flexibility of the Gaussian Process specified by $C$ will obviate
the need for numerous high order polynomial terms in $f(\alpha)$ and
thus reduce the number of free parameters required in the model.  The
prior over functions implied by $C$ is analytically conditioned on the
residuals $\Delta$.
This---equation~(\ref{eq:spectroscopicLF})---is the log-likelihood we
are going to use for the spectroscopic data in what follows.

In detail, we are going to choose the exponential of a $M$th-order
polynomial form for the calibration vector $f(\alpha)$ and a bell
shape (Gaussian) for the kernel function $K_\alpha(\cdot)$:
\begin{eqnarray}\displaystyle
f(\alpha) &=& \exp\left(\sum_{m=o}^M c_m\,x^m\right)
\\
x &\equiv& \frac{\lambda}{\lambda_0} - 1
\\
K_\alpha(\lambda_n - \lambda_{n'}) &=& \mu_n\,\mu_{n'}\, a^2\,\exp\left(-\frac{[\lambda_n - \lambda_{n'}]^2}{2\,\ell^2}\right)
\\
\transpose{\alpha} &\equiv& \left[ \{c_m\}_{m=0}^M, s^2, a^2, \ell^2 \right]
\quad ,
\end{eqnarray}
where $\lambda$ is a column vector made up of the wavelengths
$\lambda_m$ and $x$ is a scaled and shifted version of the same,
scaled by a fiducial wavelength $\lambda_0$.  The exponentiation of
the polynomial in ensures that $f(\alpha)$ is everywhere positive.
Nuisance parameters $c_m$ set the shape of the large-scale calibration
issues, while $a^2$ and $\ell^2$ set the variance and correlation
length of the remaining covariant calibration noise. The
multiplication by $\mu_n\, \mu_{n'}$ in the kernel function signifies
signifies that it is defined in \emph{fractional} terms.

@@

\subsection{Star Cluster Model}
\label{sec:cluster_model}
We model the stellar cluster as a single age SSP. The spectrum of a
star cluster under this assumption is given by a vector
\begin{eqnarray} \label{eq:StarModel}\displaystyle
f_{\theta_s}(\lambda_o) & = &
L(\sigma_s, z) ,\, \frac{m_*}{4\pi d_L^2(1+z)} \, s(\lambda, t, Z , \beta) \, e^ {-k_\lambda\tau_V} \\
\nonumber \\ 
\lambda_o & = & (1+z)\,\lambda 
\\
\theta_s & \equiv & \left[ m_*, t, Z, \tau_V, z, d_L, \sigma_*, \beta \right]
\quad ,
\end{eqnarray} where the
parameters $\theta_s$ include 
the stellar mass $m_*$, 
effective optical depth to dust in the $V$-band $\tau_V$, 
luminosity distance $d_L$,
and redshift $z$.
Multiplication by the matrix $L(\sigma_s, z)$ represents a convolution
with the stellar velocity profile,
and $\sigma_s$ is the stellar velocity dispersion. 
The shape of the dust attenuation curve is given by $k_\lambda \equiv
\tau_\lambda/\tau_V$.  
The observed frame wavelengths $\lambda_o$ are related to the
restframe wavelengths $\lambda$ in the usual way.

@@

\item Mock uncalibrated data:  vary the photometric bands to see how
  much it changes the recovery of the parameters.

@@

Next we will \emph{not} assume that the calibration is perfectly known
and we will therefore allow the calibration parameters $\alpha$ in our
model to vary.  {\color{red} This is not converged.} In Figure
\ref{fig:speconly_calibrated} we show the result of our inference from
the spectrum and a single photometric measurement (SDSS $g$) for a
perfectly calibrated mock spectrum. To demonstrate the quality of the
fit we show both the mock input spectrum and predictions of the
observed spectra that result from several draws of the model
parameters $[\theta, \alpha]$ from posterior distribution.  We also
show samples of the inferred calibration vector,
\begin{eqnarray} \label{eq:calibration}
\frac{d}{\mu(\theta)} & = & f(\alpha) + \frac{\tilde{\Delta}}{\mu(\theta)}
\quad ,
\end{eqnarray}
where $\tilde{\Delta}$ is the Gaussian Process mean prediction for the
logarithmic residual between the data and the model \citep{RW}.  For the
perfectly calibrated spectrum the true calibration vector (shown in
black) is one everywhere.  The spread of the inferred calibration
vectors approximately indicates our uncertainty in the spectrophotometric
calibration when no additional calibration information is available
and only a single photometric point is available.

@@

In Figure \ref{fig:speconly_uncalibrated} we show the results of
inference on an \emph{uncalibrated} mock spectrum.  The results are
similar to the case of the \emph{calibrated} spectrum.

@@

We have demonstrated, using using both mock and real spectra of
galactic stellar clusters, that we can combine photometric and
spectroscopic data in a principled way to infer physical properties.
We do not resort to an arbitrary down-weighting or up-weighting of
certain data. Instead, we explicitly and flexibly model our uncertain
knowledge of the spectral calibration and infer this calibration
jointly with the physical properties. For a number of applications
this combination of spectroscopy and photometry reduces the need for
spectrophotometric calibration entirely, such that the less expensive
photometric calibration is sufficient.  Additional calibration
information, if it exists, can be incorporated into informative priors
on the calibration parameters of the model, inform the construction of
the calibration model (including the form of the covariance kernel),
or approximately remove gross features that are difficult to model
parametrically or via a Gaussian Process.  {\color{blue} Discuss more
complicated covariance kernels, multiple kernels, combining fractional
and absolute covariance amplitudes.}

@@

Limitations.  The strategy we have advocated has several limitations.
First, it requires a generative model for the spectrum of the object
of interest, and is thus ill-suited to discovery projects.  Second,
care must be taken to avoid absorbing interesting model deficiencies
into the calibration model. For example, in the model we have
described, it is possible that broad shallow features (for example due
to molecular bands or absorption line complexes) will be absorbed into
the Gaussian Process if not properly included in the mean model. The
Gaussian process will also absorb incorrectly modeled attenuation
curve shapes. Third, we have made the assumption that the spectroscopy
and the photometry are sampling the same populations.  In practice,
differences between spectroscopic and photometric apertures often mean
that the photometry and the spectroscopy are sampling slightly
different physical regions.  The correct response to this situation is
to explicitly model the differences, incorporating prior information
about the spatial variation of physical properties.  Finally, this
method is computationally intensive and expensive, since we are
constantly inverting largish matrices.

>>>>

A model is a parameterized likelihood function, with prior PDFs
on the nuisance parameters (at least).  There is a likelihood for
the photometry and for the spectroscopy.  Each of these penalizes
the difference between a column vector of data and a column vector of
model expectation value.

@@

If you don't believe the reported data noise variances $\sigma_n^2$,
it often makes sense to add in quadrature a ``jitter'' variance $s^2$.
This is a simple modification of the chi-squared log-likelihood:
\begin{eqnarray}\label{eq:photometricLF}
\ln p(d\given\theta,s^2) &=& -\frac{1}{2}\,\sum_{n=1}^N
                             \left[\frac{[d_n - \mu_n(\theta)]^2}
                             {[\sigma_n^2 + s^2]} + 
                             \ln(2\pi\,[\sigma_n^2 + s^2]) \right]
\quad .
\end{eqnarray}
This modification shows the necessity of including the log-variance
term.  Large $s$ reduces the penalty for individual large residuals,
but also reduces the reward for well-fit data.
This---equation~(\ref{eq:photometricLF})---is in fact the
log-likelihood function we are going to use for the photometric
measurements in what follows.  For photometric data this ``jitter'' is
often best described as a fractional error, $s^2 \equiv
\hat{s}^2\mu(\theta)^2$, corresponding for example to an additional
magnitude zero-point uncertainty.  In this formulation, if $\theta$
are the parameters of interest, the jitter variance $\hat{s} =
s/\mu(\theta)$ is a nuisance noise-model parameter representing
additonal fractional uncertainty in the flux.

{\color{blue} BDJ This paragraph should explain why we are using the
GP, not just how it works.}
For the spectroscopic measurements we are going to account for the
spectrophotometric calibration in our likelihood function.  We take
the calibration-vector that modifies the mean spectroscopic model to
produce the observed spectrum to be multiplicative with respect to the
mean model.  The multiplicative calibration vector can be explicitly
modeled, for example using polynomials.  However, for complex
calibration vectors over large wavelength ranges, the number of
parameters required to model the calibration can become quite large
(e.g. very high order polynomials) and the correct parameterization
may be unknown.  It is therefore desirable to introduce a
`non-parameteric' model of the calibration.  We do this using a
Gaussian Process in the likelihood function.  The Gaussian Process is
a generalization of the modified log-likelihood introduced in
Equation \label{eq:photometricLF} that allows for correlated noise in
the residuals, as would be introduced by inaccurate calibration.

%BDJ: prior over functions, analytically marginalizable, covariant
%fractional uncertainties the same as multiplying by a function
%(covariant absolute uncertainties the same as adding a function.)

The calibration-vector and Gaussian-Process modified log-likelihood is
\begin{eqnarray}\label{eq:spectroscopicLF_v2}
\ln p(d\given\theta,\alpha) &=&
                                -\frac{1}{2}\,\left[\transpose{\Delta}\,C^{-1}\,\Delta
                                + \ln([2\pi]^N\,\det{C}) \right]
\\
\Delta &\equiv& d - \mu(\theta) \ast f(\alpha) 
\\
C_{nn'} &\equiv& [\sigma_n^2 + (\hat{s} \, \mu_n)^2]\,\delta_{nn'} +
 K_\alpha(\lambda_n - \lambda_{n'}) 
\quad ,
\end{eqnarray}
where $\ast$ denotes element-wise multiplication,
$\theta$ are the physical parameters,
$\alpha$ is a blob of nuisance parameters
(which determine the jitter, the calibration vector and the kernel function),
$\Delta$ is a residual column vector,
$f(\alpha)$ is a calibration column vector,
$C$ is the Gaussian Process covariance matrix,
made up of $N^2$ values $C_{nn'}$,
$K_\alpha$ is a kernel function that sets the wavelength scale and
amplitude of the residual calibration issues,
and each $\lambda_n$ is the wavelength of spectral pixel $n$.
The calibration vector $f(\alpha)$ is going to take care of the
large-scale calibration residual; the Gaussian Process is going to
pick up smaller-scale residuals away from \emph{that}.  The
flexibility of the Gaussian Process specified by $C$ will obviate the
need for numerous high order polynomial terms in $f(\alpha)$ and thus
reduce the number of free parameters required in the model.
This---equation~(\ref{eq:spectroscopicLF})---is the log-likelihood we
are going to use for the spectroscopic data in what follows.

In detail, we are going to choose the exponential of a $M$th-order
polynomial form for the calibration vector $f(\alpha)$ and a bell
shape (Gaussian) for the kernel function $K_\alpha(\cdot)$:
\begin{eqnarray}\displaystyle
f(\alpha) &=& \exp\left(\sum_{m=o}^M c_m\,x^m\right)
\\
x &\equiv& \frac{\lambda}{\lambda_0} - 1
\\
K_\alpha(\lambda_n - \lambda_{n'}) &=& \mu_n\,\mu_{n'}\, a^2\,\exp\left(-\frac{[\lambda_n - \lambda_{n'}]^2}{2\,\ell^2}\right)
\\
\transpose{\alpha} &\equiv& \left[ \{c_m\}_{m=0}^M, s^2, a^2, \ell^2 \right]
\quad ,
\end{eqnarray}
where $\lambda$ is a column vector made up of the wavelengths
$\lambda_m$ and $x$ is a scaled and shifted version of the same,
scaled by a fiducial wavelength $\lambda_0$.  The exponentiation of
the polynomial ensures that $f(\alpha)$ is everywhere positive.
Nuisance parameters $c_m$ set the shape of the large-scale calibration
issues, while $a^2$ and $\ell^2$ set the variance and correlation
length of the remaining covariant calibration noise. Structure in the
residuals on the scale of $\ell$ and with amplitude $a$ will not be
strongly penalized, but the determinant of $C$ becomes larger for
larger $a$ and smaller $\ell$. The multiplication by $\mu_n\,
\mu_{n'}$ in the kernel function signifies that it is defined in
\emph{fractional} terms.  Using a high-order polynomial can result in
strong degeneracies between model and polynomial parameters, whereas
the GP {\color{blue} BDJ make this more precise and correct} only
reduces the likelihood penalty for residuals on the scale of $\ell$
with amplitude $a$ that {\it can't} be fit by the mean model.

@@

\subsection{Star Cluster Model}
\label{sec:cluster_model}
We model the stellar cluster as a single age SSP. The spectrum of a
star cluster under this assumption is given by a vector
\begin{eqnarray} \label{eq:StarModel}\displaystyle
f_{\theta_s}(\lambda_o) & = &
L(\sigma_s, z) \, \frac{m_*}{4\pi d_L^2(1+z)} \, s(\lambda, t, Z , \beta) \, e^ {-k_\lambda\tau_V} \\
\nonumber \\ 
\lambda_o & = & (1+z)\,\lambda 
\\
\theta_s & \equiv & \left[ m_*, t, Z, \tau_V, z, d_L, \sigma_*, \beta \right]
\quad ,
\end{eqnarray} where the
parameters $\theta_s$ include 
the stellar mass $m_*$, 
effective optical depth to dust in the $V$-band $\tau_V$, 
luminosity distance $d_L$,
and redshift $z$.
Multiplication by the matrix $L(\sigma_s, z)$ represents a convolution
with the stellar velocity profile,
and $\sigma_s$ is the stellar velocity dispersion. 
The shape of the dust attenuation curve is given by $k_\lambda \equiv
\tau_\lambda/\tau_V$.  
The observed frame wavelengths $\lambda_o$ are related to the
restframe wavelengths $\lambda$ in the usual way.

@@

\item \sout{Mock uncalibrated data:  vary the photometric bands to see how
  much it changes the recovery of the parameters.}

@@

Next we will \emph{not} assume that the calibration is perfectly known
and we will therefore allow the calibration parameters $\alpha$ in our
model to vary.  {\color{red} This is not converged.} In Figure
\ref{fig:speconly_calibrated} we show the result of our inference from
the spectrum and a single photometric measurement (SDSS $g$) for a
perfectly calibrated mock spectrum. To demonstrate the quality of the
fit we show both the mock input spectrum and predictions of the
observed spectra that result from several draws of the model
parameters $[\theta, \alpha]$ from posterior distribution.  We also
show samples of the inferred calibration vector,
\begin{eqnarray} \label{eq:calibration}
\frac{d}{\mu(\theta)} & = & f(\alpha) + \frac{\tilde{\Delta}}{\mu(\theta)}
\quad ,
\end{eqnarray}
where $\tilde{\Delta}$ is the Gaussian Process mean prediction for the
residual between the data and the model \citep{RW},
defined here as 
\begin{eqnarray} \label{eq:calibration}
\tilde{\Delta} & = & K \, C^{-1} \, \Delta
\end{eqnarray}.  
For the perfectly calibrated spectrum the true calibration vector
(shown in black) is one everywhere.  The spread of the inferred
calibration vectors approximately indicates our uncertainty in the
spectrophotometric calibration when no additional calibration
information is available and only a single photometric point is
available.

@@

In Figure \ref{fig:speconly_uncalibrated} we show the results of
inference on an \emph{uncalibrated} mock spectrum.  The results are
similar to the case of the \emph{calibrated} spectrum. An important
difference is that the inferred calibration vector now contains numerous small scale wiggles,
similar to the actual calibration vector.  The overall shape of the
spectrum is still unconstrained, leading to large degeneracies between
the polynomial terms and the dust attenuation.

@@

We have demonstrated, using using both mock and real spectra of
galactic stellar clusters, that we can combine photometric and
spectroscopic data in a principled way to infer physical properties.
We do not resort to an arbitrary down-weighting or up-weighting of
certain data. Instead, we explicitly and flexibly model our uncertain
knowledge of the spectral calibration and infer this calibration
jointly with the physical properties. For a number of applications
this combination of spectroscopy and photometry reduces the need for
spectrophotometric calibration entirely, such that the less expensive
photometric calibration is sufficient.  Additional calibration
information, if it exists, can be incorporated into informative priors
on the calibration parameters of the model, inform the construction of
the calibration model (including the form of the covariance kernel),
or approximately remove gross features that are difficult to model
parametrically or via a Gaussian Process. A Guassian Process can be
used to model covaraint noise resulting from errors in an initial
calibration (of an amplitude $a$) thus propogating calibration
uncertainty into the model parameters $\theta$. 
{\color{blue} Discuss more complicated covariance kernels, multiple
kernels, combining fractional and absolute covariance amplitudes.}

@@

Limitations.  The strategy we have advocated has several limitations.
First, it requires a generative model for the spectrum of the object
of interest, and is thus ill-suited to discovery projects.  Second,
care must be taken to avoid absorbing interesting model deficiencies
into the calibration model, or interpreting features in the
calibration vector as features of the mean model. For example, in the
model we have described, it is possible that broad shallow features
(for example due to molecular bands or absorption line complexes) will
be absorbed into the Gaussian Process if not properly included in the
mean model. The Gaussian process will also absorb incorrectly modeled
attenuation curve shapes. On the other hand, the higher likelihood at
zero residual means that calibration features that \emph{can} be
mimicked by the mean model may influence the mean model
parameters. Third, we have made the assumption that the spectroscopy
and the photometry are sampling the same populations.  In practice,
differences between spectroscopic and photometric apertures often mean
that the photometry and the spectroscopy are sampling slightly
different physical regions.  The correct response to this situation is
to explicitly model the differences, incorporating prior information
about the spatial variation of physical properties.  Finally, this
method is computationally intensive and expensive, since we are
constantly inverting largish matrices.
